Benchmark!!1

### 2021-02-17 paper/main.tex
paper progress

Philosophy: not necessarily the best
implementation in the world, but a reasonably good one, i.e. one we
can expect from an average student. Similar algorithm in various
languages. Use as few external libraries as possible, but use them if
needed.

@@

In order to represent a tour in heuristics we use variable-size
vectors, since they are necessary in most algorithms for 
vehicle routing, for instance to insert or delete a vertex from a
tour.
% This is the reason why we do
% not provide a C implementation. It is of course possible to
% implement variable-size vectors using fixed-size arrays and
% bookkeeping; however this stands directly against our stated purposes
% and philosophy. Moreover, implementing a bug-free, efficient version
% of variable-size vectors is unlikely to result in something better
% than existing libraries.
Inserting and deleting elements in a vector has linear worst-case time
complexity. However, for heuristics it is still usually a good idea to
use vectors rather than, say, linked lists, since neighborhood
exploration typically requires to evaluate many moves then perform
only one of these moves. The complexity of performing a move is thus
secondary and unlikely to be a performance bottleneck. Additionally,
performing certain moves can actually be more efficient using a
vector than using a linked list, e.g. inverting a subsequence.

@@

\section{Languages and implementations considered}
We consider six different programming languages: C++, Python, Java,
Julia, Rust and Javascript. Any rule of selection
is arbitrary in nature; however, we indicate below, for each of these
languages, the main reasons behind their selection. It is possible to
add other languages to this list in the future.

@@

\subsection{C++}
Two implementations: c++14 vs c++98.

Flat vs nested matrix.

static arrays too.

\subsection{Python}
Python vs Pypy vs Numpy vs Numba.
Flat vs nested matrix.

\subsection{Java}
Flat vs nested matrix.
static arrays too.

\subsection{Julia}
General performance tips: read the specific doc section! One unique
official document outlines how to write efficient code. there is also
an automated tool for that.

Flat vs nested matrix.

@@

\subsection{Other general cross-language considerations}
Here we discuss considerations that are subjective in nature, as they
relate to our experience implementing all benchmarks in all
languages. We need to take into account our experience using various
languages prior to the beginning of this work:
\begin{itemize}
\item C++: numerous projects over 20 years.
\item Python: numerous projects over 15 years; no experience of Numpy
  or Numba.
\item Java: PhD work 15 years ago, one minor project 5 years ago.
\item Julia: one ongoing research project started 3 years ago.
\item Javascript: a few hobby projects 5-7 years ago.
\item Rust: no experience at all.
\end{itemize}
To add to the subjectivity, each benchmark was first implemented in
Python then in other languages, with Rust typically being the last one
to be implemented. Therefore the languages implemented last were
supposedly easier, due to increased familiarity with the
algorithm. For all reasons mentioned above, measuring the
time spent implementing each benchmark in each language would be
meaningless, therefore we limit ourselves to subjective opinion here.
Keeping this in mind, we see a subset of perceived
\emph{easy} languages, as in ``easy to program in and get things running
with the expected outcome'', which consists of Python, Julia and
JavaScript. These 
easy languages typically allow to program the same algorithms faster
than the other ones with the same results, performance considerations
aside.
On the other hand, we found the process of getting benchmarks to work
with Rust to be much more tedious than with any other language,
although this is certainly due to lack of experience. Nonetheless, the
JavaScript implementation was typically easier than the C++ or Java ones
despite having more experience with C++ and Java, so
experience does not explain everything.
Ultimately, there is little doubt that any experienced programmer in
any of these languages would have no trouble implementing any of these
benchmarks. Nonetheless, this programmer with experience in mostly C++
and Python implemented every benchmark faster in Python, JavaScript or Julia
than in C++.

@@

Using two threads instead of 1 improves performance by about 10\%,
although it is unclear what the second thread is used for.

@@

>>>>

\subsection{Scope of this study}
We want to compare the reasonably efficient implementation of various
algorithms used in combinatorial optimization. In particular, we look
at five algorithms that are commonly used in vehicle routing
optimization. However these algorithms have features that are also
present in other domains of optimization, such as modifying
subsequences, inserting and removing elements in a sequence, or other
operations on graphs.

Our goal is not to provide the most efficient implementation of each
algorithm for a set of languages, or even to use the most efficient
algorithm to solve the given problems, but rather to use these
algorithms as benchmarks, each capturing certain features currently
encountered in optimization. We put ourselves in the situation of an
average student having to implement these algorithms, and compare the
performance of different programming languages in that context. We
also assess the impact on performance of some implementation
decisions, e.g. using a flat matrix representation.

@@

In order to represent a tour in heuristics we use variable-size
vectors. Dynamic size containers are necessary in most algorithms for 
vehicle routing, for instance to insert or delete a vertex from a
tour.
% This is the reason why we do
% not provide a C implementation. It is of course possible to
% implement variable-size vectors using fixed-size arrays and
% bookkeeping; however this stands directly against our stated purposes
% and philosophy. Moreover, implementing a bug-free, efficient version
% of variable-size vectors is unlikely to result in something better
% than existing libraries.
Inserting and deleting elements in a vector has linear worst-case time
complexity. However, for heuristics it is still usually a good idea to
use vectors rather than, say, linked lists, since neighborhood
exploration typically requires to evaluate many moves then perform
only one of these moves. The complexity of performing a move is thus
secondary and unlikely to be a performance bottleneck. Additionally,
performing certain moves can actually be more efficient using a
vector than using a linked list, e.g. inverting a subsequence in the
context of 2-opt.

@@

\section{Languages and implementations considered}
We consider six different programming languages: C++, Python, Java,
Julia, Rust and Javascript. Any rule of selection
is arbitrary in nature; however we indicate below, for each of these
languages, our perceived advantages of these languages, which justify
their selection. It is possible to add other languages to this list in
the future, and keep the benchmark collection alive in an online form.
For a given language several implementations are possible, e.g. one
with nested distance matrices and one with flat distance matrices. As
a convention, implementations are named in lower-case letters, for
example \emph{python-flat-matrix}. We also use the simplest names for
the best implementations, i.e. \emph{python} uses a nested distance
matrix because it performs better than \emph{python-flat-matrix}, but
\emph{pypy} uses a flat matrix because it performs better than
\emph{pypy-nested-matrix}. All implementations are summarised in
Appendix~\ref{app:impl}.

@@

It is difficult to properly measure the popularity of a programming
language. There exist rankings, for example based on online
searches~\cite{TIOBE} or on GitHub contributions~\cite{octoverse}. All
these approaches have flaws. Basing an index on online searches is
subject to all the biases and exploits that target online
search. Measuring GitHub contributions is not clear, as
there is no trivial way to decide what should count among
repositories, number of lines written, number of commits,
etc. Ultimately, language popularity for its own sake does not need to
be a crucial factor in the choice of a programming language, especially for
academic research. These aspects seem more important to us:
\begin{itemize}
\item The programs written in a given programming language should be
  easily run on any computer. For that reason, we only look at
  languages that are free and work under multiple operating
  systems. The ability to run under Linux is especially important,
  since it is typically the operating system of high-performance
  computing (HPC) clusters~\cite{top500}.
\item The language should be well documented, but also have online
  resources for support. This in fact also depends on popularity.
\end{itemize}

\subsection{C++}
C++ is one of the most popular programming languages in the World, and
has been so for decades. Among its many features there is good
runtime performance. In fact the performance is good enough that whole
operating systems can be written in C++, for example Microsoft
Windows. In fact C++ has so many features that it is often possible to
implement the same algorithm using different subsets of the language,
and a frequent criticism is that people and teams evolve into using
only their own subset of the language.

We develop two C++ implementations of every benchmark, using the C++98
and C++14 standards, which we call \emph{c++98} and \emph{c++14}. they
both use a flat distance matrix. With C++98 we use references to
objects, while with C++14 we use smart pointers such as
\verb|shared_ptr|. Comparing these two implementations will allow us
to evaluate the cost of smart pointers.

For each benchmark the reference CPU time is provided by the c++14
implementation. We also develop a variant of c++14 using a nested
distance matrix, and another using static arrays.

\subsection{Python}
Python is another of the most popular programming languages in the
World, especially for tasks of machine learning in recent years. The
base Python implementation is called \emph{python} and uses a nested
distance matrix. There is also an implementation using a flat distance
matrix, called \emph{python-flat-matrix}, and one using a nested
distance matrix wrapped in function calls, called
\emph{python-nested-matrix-function}.

One issue with Python is that much of the code is not compiled, rather
directly interpreted, which can lead to poor performance. One way to
address that issue is to rely on libraries that are implemented in C,
however that is not always possible. We consider alternative
implementations in Python, all bringing different solutions to this
issue.

The simplest way to speed up the run time of a Python program is
to run this program with Pypy~\cite{pypy} instead of the
standard CPython interpreter. Pypy uses JIT compilation to speed
things up. There is no need to modify the Python program, which is a
strong advantage compared to library-based solutions or
annotation-based solutions. Pypy's website claims that it is on
average 4.2 times faster than CPython~\cite{pypy}, however this is
just an average and actual numbers may vary a lot depending on the
benchmark. We use both the python and python-flat-matrix
implementations with Pypy, resulting in \emph{pypy-nested-matrix} and
\emph{pypy} implementations, respectively.

We also implement the benchmarks using Numpy arrays instead
of standard Python lists. Numpy is one of the most popular external
libraries for Python; it is coded in C for runtime
performance~\cite{numpy}. Numpy performs especially well when using
\emph{array operations}, for instance matrix multiplication. However
the benchmarks that we consider do not take advantage of Numpy
functions, only of its array data structure. Since Numpy provides
multi-dimensional arrays, we use a two-dimensional array to represent
the distancematrix. Additionally, we consider a flat matrix in
implementation \emph{numpy-flat-matrix}.

Numpy itself does not provide any kind of JIT compilation, but another
Python package, Numba, performs JIT compilation when using Numpy
arrays. Numba requires code annotation, therefore we annotate the
Numpy implementations to take advantage of them with Numba.

\subsection{Java}
Java is a very popular programming language of the past two
decades. Java code is compiled to \emph{bytecode}, which is then run
by a virtual machine (VM). The bytecode compiled on one computer can
be run by a VM on another computer. Nonetheless, JIT compilation to
native code happens dynamically at run time. The syntax is based on
C/C++ syntax. The base Java implementation, \emph{java}, uses a flat
distance matrix and a \verb|ArrayList<Integer>| object to represent a
tour. We also implement a version with nested distance matrix,
\emph{java-nested-matrix}, as well as a version with static arrays to
represent a tour, \emph{java-static-arrays}.

\subsection{Julia}
Julia is a relatively recent programming language. Mention dates for
first release (0.4?) as well as 1.0.

Flat vs square matrix, mention 2-dimensional arrays.

@@

\subsection{Other general cross-language considerations}
Here we discuss considerations that are subjective in nature, as they
relate to individual experience implementing all benchmarks in all
languages. Experience using various languages prior to the beginning
of this work needs to be taken into account:
\begin{itemize}
\item C++: numerous projects over 20 years.
\item Python: numerous projects over 15 years; no experience of Numpy
  or Numba.
\item Java: PhD work 15 years ago, one minor project 5 years ago.
\item Julia: one ongoing research project started 3 years ago.
\item Javascript: a few hobby projects 5-7 years ago.
\item Rust: no experience at all.
\end{itemize}
To add to the subjectivity, each benchmark is first implemented in
Python then in other languages, with Rust typically being the last one
to be implemented. Therefore the languages implemented last represent
supposedly easier tasks, due to increased familiarity with the
algorithm. For all reasons mentioned above, measuring the
time spent implementing each benchmark in each language would be
meaningless, therefore we limit ourselves to subjective opinion here.
Keeping this in mind, we see a subset of perceived
\emph{easy} languages, as in ``easy to program in and get things running
with the expected outcome'', which consists of Python, Julia and
JavaScript. These 
easy languages typically allow to program the same algorithms faster
than the other ones with the same results, performance considerations
aside.
On the other hand, we found the process of getting benchmarks to work
with Rust to be much more tedious than with any other language,
although this is certainly due to lack of experience. Nonetheless, the
JavaScript implementation was typically easier than the C++ or Java ones
despite having more experience with C++ and Java, so
experience does not explain everything.
Ultimately, there is little doubt that any experienced programmer in
any of these languages would have no trouble implementing any of these
benchmarks. Nonetheless, this programmer with experience in mostly C++
and Python implemented every benchmark faster in Python, JavaScript or Julia
than in C++.

@@

% Using two threads instead of 1 improves performance by about 10\%,
% although it is unclear what the second thread is used for.

@@

\appendix

\section{Summary of implementations}
\label{app:impl}
\begin{table}
  \begin{tabular}{l|llll}
    \toprule
    Implementation & Language & Extra & Distance matrix representation
    & Note \\
    \midrule
    \bottomrule   
  \end{tabular}
  \caption{Overview of all implementations}
\end{table}
